---
title: "ENVS193ds-final"
author: "Valeria Fierros"
date: "due: 2026-03-17"
format: 
  html: 
    toc: true
execute: 
  message: false
  warning: false
editor: visual
---

[Final GitHub Repo](https://github.com/valeriafierros/ENVS-193DS_winter-2026_final.git)

# Set up

```{r}
#| label: set up

#reading in packages here
library(tidyverse)
library(janitor)
library(ggplot2)
library(MuMIn)
library(DHARMa)
library(ggeffects)
library(effsize)

#reading in data here
kelp <- read.csv("../data/kelp_data.csv")
mopl <- read.csv("../data/mopl.csv")
steps <- read.csv("../data/cyopersonaldata - Sheet1.csv")
```

# Problem 1: Research writing

## 1a. Transparent statistical methods

In part 1, they used a simple linear regression to determine whether precipitation predicts wetland flooding. In part 2, they used a Kruskal-Wallis test to determine whether median flooded wetland area differs across various levels of water year classifications.

## 1b. Figure needed 

They can use a boxplot to visualize these results. Because there are multiple predictors/levels, the x-axis should show the water year classification categories (wet, above normal, below normal, dry, critical drought). The y-axis should show the response variable, which would be flooded wetland area (m²). Each category on the x-axis would have its own boxplot to display the interquartile range, median, and spread of the data.

## 1c. Suggestions for rewriting

(1) Precipitation was not a strong predictor of flooded wetland area in the San Joaquin River Delta (simple linear regression, F(df model, df error) = test statistic, $R^2$ = effect size, p = 0.11, $\alpha$ = significance level).

(2) There is no significant difference in median flooded wetland area between water year classifications (wet, above normal, below normal, dry, critical drought)(Kruskal-Wallis test, $X^2$(4) = test statistic, p = 0.12, $\alpha$ = significance level).

## 1d. Interpretation

Another variable that could influence the flooded area of a wetland is local wetland management practices. This variable would be categorical (managed vs unmanaged) or ordinal (levels of management/intervention). It could influence flooding because management decisions can directly influence water flow and retention in wetlands. For example, higher water retention due to lack of vegetation in wetlands could increase flooding.

# Problem 2. Data visualization

## 2a. Cleaning and summarizing

```{r}
#| label: cleaning data

#cleaning data for problem 2
kelp_clean <- kelp |> #creating new clean kelp object from kelp df
  clean_names() |> # function 3: cleaning column names
  filter(site %in% c("IVEE", "NAPL", "CARP")) |> # function 6: filtering site column to only include Isla Vista, Naples, and Carpinteria
  filter(!(date =="2000-09-22" & transect == "6" & quad == "40")) |> # function 1: filtering out frond observations from the date `2000-09-22` in transect 6 in quad 40 
  mutate(site_name = case_when( # function 8: creating new column called site_name and turning site codes into the full names
    site == "NAPL" ~"Naples",
    site == "IVEE" ~ "Isla Vista",
    site == "CARP" ~ "Carpinteria")) |> 
  mutate(site_name = as_factor(site_name), # function 2: making site_name a factor with 3 levels (Naples, Isla Vista, and Carpinteria)
         site_name = fct_relevel(site_name,
                                 "Naples", "Isla Vista", "Carpinteria")) |> 
  group_by(site_name, year) |> # function 5: grouping by site_name and year
  summarize(mean_fronds = mean(fronds)) |> # function 7: calculating mean frond length
  ungroup() # function 4: ungrouping site_name and year

#checking data
slice_sample(kelp_clean,
             n = 5) #displaying 5 random rows from kelp_clean data frame
str(kelp_clean) # displaying kelp_clean data frame structure
```

## 2b. Understanding data

This line filters out frond observations from this date in transect 6 in quad 4 because there is a missing value there; they are input as -99999, which represents no observation. This information can be found on the data page under the files and data links tab under FRONDS.

## 2c. Visualize the data

```{r}
#| label: recreating visualization
#| fig-height: 5
#| fig-width: 8

#base layer
ggplot(data = kelp_clean, #using kelp_clean data
       aes(x = year, #setting x axis
           y = mean_fronds, #setting y axis
           color = site_name))+ #color by each site 
  geom_line(size = 0.3)+ #connecting points with a line and setting line size
  geom_point()+ # adding individual data points
  scale_color_manual(values = c( # setting manual colors
    "Naples" = "cadetblue", #naples color
    "Isla Vista" = "coral", #IV color
    "Carpinteria" = "blue4" #carp color
  ))+
  labs(title = "Sites vary in mean kelp fronds per year", #adding title
       x = "Year", #labelling x axis
       y = "Mean kelp fronds per year", #labelling y axis
       color = "Site")+ #renaming legend
  theme_minimal()+ #changing theme
  theme(plot.title = element_text(hjust = 0.5), #centering the title
        text = element_text(family = "sans"), # using custom font
        panel.border = element_blank()) #making panel background blank


```

## 2d. Interpretation

Isla vista has the most variable mean kelp fronds per year; its line shows the largest fluctuations over time with the strongest peaks and dips. The maximum points goes all the way up to \~ 40 while the minimum dips down to less than five. The site with the least variable mean kelp fronds per year is Carpinteria. The points from this site have less variation and a smaller range, with less dramatic peaks occurring. Each observation remains between the same range (\~5-20).

# Problem 3. Data analysis

## 3a. Response variable

The response variable is Mountain Plover occupancy. In this data set, the 1s and 0s are binary numbers that represent mountain plover nest presence/absence at a given nesting site when surveyed. A 1 means the species nest was observed while a 0 means there was no nests present at the time of the survey.

## 3b. Purpose of study

**Visual obstruction**

Visual obstruction is a continuous variable where each observation is an individual measurement, explained in the data collection subsection of the methods section in paragraph 3. As visual obstruction increases, the probability of Mountain Plover nest site use decreases because Mountain Plovers tend to prefer flat landscapes with bare ground exposure and sparse vegetation (found in paragraph 2 of the introduction).

**Distance to prairie dog colony edge**

Distance to prairie dog colony edge is also a continuous variable, found in the modeling framework subsection of the methods section in paragraph 1. As distance to colony edge increases, the probability of Mountain Plover nest site decreases because vegetation height and shrub cover are higher outside of prairie dog colonies where Mountain Plovers prefer to nest (found in the 3rd paragraph of the discussion).

## 3c. Table of models

| Model number | Visual Obstruction | Distance to Colony | Model Description |
|------------------|:-----------------:|:-----------------:|:----------------:|
| 0 |  |  | no predictors( null model) |
| 1 | X | X | all predictors (saturated model) |
| 2 |  | X | distance to prairie dog colony only |
| 3 | X |  | visual obstruction only |

## 3d. Run the models

```{r}
#| label: running models

#cleaning data
mopl_clean <- mopl |> # creating new mopl_clean object from mopl data frame
  clean_names() |> # cleaning column names
  mutate(used_bin = case_match(id, # creating new used_bin column and creating binomial variables where used = 1 and available = 0
                               "used" ~ 1,
                               "available" ~ 0)) |> 
  select(used_bin, edgedist, vo_nest) #selecting 3 desired columns

#model 0 - no predictors
mod0 <- glm(used_bin ~ 1, #running glm with no predictors
            data = mopl_clean, #using cleaned data frame
            family = "binomial") #setting binomial distribution

#model 1 - all predictors
mod1 <- glm(used_bin ~ edgedist + vo_nest, #running glm with both predictors (visual obstructions and edge distance)
            data = mopl_clean, # using cleaned data frame
            family = "binomial") #setting binomial distribution

#model 2 - only prairie distance
mod2 <- glm(used_bin ~ edgedist, # running glm with only edge distance as a predictor
            data = mopl_clean, #using cleaned data frame
            family = "binomial") #setting binomial distribution

#model 3 - only visual obstruction
mod3 <- glm(used_bin ~ vo_nest, #running glm with only visual obstructions as a predictor
            data = mopl_clean, #using cleaned data frame
            family = "binomial") #setting binomial distribution
```

## 3e. Select the best model

```{r}
#| label: running AICc

AICc(mod0, mod1, mod2, mod3) |> #running AIC for all 4 models
  arrange(AICc) # arranging model AICc from lowest to highest scores

```

The best model was determined by Akaike's Information Criterion (AIC) to be model 3, which has a score of 331.81. This model only has visual obstruction as its predictor variable and Mountain Plover nest presence/use as its response variable; therefore, Mountain Plover nesting is best predicted by visual obstructions.

## 3f. Check the diagnostics

```{r}
#| label: checking diagnostics
#| fig-height: 6
#| fig-width: 8

plot(simulateResiduals(mod3)) #plotting QQ plot and residual vs predicted with DHARMa using model 3

```

## 3g. Visualize the model predictions

```{r}
#| label: visualizing model prediction

mod_preds <- ggpredict(mod3,#creating new prediction object using model 3
                       terms = "vo_nest") #using visual observation as predictor

#creating ggplot: base layer
ggplot(mopl_clean, #using clean mopl data frame
       aes(x = vo_nest, #setting x acis
           y = used_bin))+ #setting y axis
  geom_point(size = 3, #adding data points and setting size
             alpha = 0.5, #transparency of data points
             color = "hotpink2")+ #making points pink
  geom_line(data = mod_preds, #adding trendline of predictions
            aes(x = x, #x axis
                y = predicted, #y axis
                color = "red"))+ #making trendline red
  geom_ribbon(data = mod_preds, #adding confidence interval from model prediction
              aes(x = x, #x axis
                  y = predicted, #y axis
                  ymin = conf.low, #lower CI range
                  ymax = conf.high), #upper CI range
              alpha = 0.3, #ribbon transparency
              fill = "pink")+ #ribbon color
  theme_minimal()+ #changing theme
  theme(panel.grid = element_blank(), #removing gridlines
        panel.background = element_blank(), #making background blank
        legend.position = "none")+ #removing legend
  labs(title = "Mod3: As visual obstructions increase, Mountain Plover nesting site use decreases", # plot title
       x = "Visual Obstruction (dm)", #labelling x axis
       y = "Nest Site Use")+ # labelling y axis
  scale_y_continuous(breaks = c(0,1)) #setting y axis interval (0 to 1)
  
  
```

## 3h. Write a caption for your figure

**Figure 1. As visual obstructions increase, Mountain Plover nesting site use decreases.** Each pink point represents an individual observation of nest presence. The red line represents the probability of a nesting site being present as visual obstructions increase, where 1 = 100% and 0 = 0%, and the pink ribbon (shaded area) around the line is the calculated 95% confidence interval. Nest site use is recorded as a binary variable with no use represented by 0 and used represented by 1. Data from: Courtney J Duchardt, Jeffrey L Beck, David J Augustine, Mountain Plover habitat selection and nest survival in relation to weather variability and spatial attributes of black-tailed prairie dog disturbance, The Condor: Ornithological Applications, Volume 122, Issue 1, 3 February 2020, duz059, https://doi.org/10.1093/condor/duz059

## 3i. Calculate model predictions

```{r}
#| label: calculating model predictions

ggpredict(mod3, #using model 3 for prediction
          terms = "vo_nest[0]") #predicting when visual obstructions equal 0
```

## 3j. Interpret your results

The predicted probability of Mountain Plover nest site occupancy at no visual obstruction (0) is 0.73, indicating that Mountain Plovers tend to use nests where there is less visual obstruction, supporting the negative relationship seen in Figure 1. This is consistent with the observations found in 3b; Mountain Plovers prefer to nest in open and sparsely vegetated habitats with less obstructions because it allows for higher visibility of predators. This is also in accordance with distance to prairie dog colony edge; the farther from the colony, the less visual obstructions and sparser the vegetation is, which means the probability of Mountain Plover nesting is greater.

# Problem 4. Affective and exploratory visualizations

## 4a. Comparing visualizations

In homework 2, I visualized my data using bar graphs and a scatterplot. In homework 3, I did the same thing, but updated the data with all of my observations. In the bar graph, the final graph showed the full range of observations; with the limited data in homework 2, it showed that Monday had the highest number of steps, but in my final visualization we see that I had the greatest number of steps on Thursdays. There was also a different pattern on my scatterplot. The initial plot showed a decrease in step count over time, but the final plot shows that there is no visible correlation between date and steps taken (non-linear). I think that this data would be better visualized by a boxplot instead, so if we were to make another visualization, that's what I would do (which is also feedback I got from my classmates during week 9 workshop). In my affective visualization, you can see the pattern/correlation between day of the week and step count better, like in the bar graph, though the bar graph uses total step count for the day and the affective visualization uses individual/daily counts. The scatterplot is the only visualization by which you can't see the pattern, and that is because my predictor variable was the date, which would not make sense in predicting step count. For a future visualization, I would like to keep day of the week as the predictor, but break it up into 2 categories – weekday and weekend –, which I think would give us a better analysis.

## 4b. Designing an analysis

I think that an appropriate analysis for this would be a two-sample t-test to test whether my daily step count differs between weekdays vs weekends. This analysis would be suitable because step count is continuous and the predictor (day type) is categorical.

## 4c. Check any assumptions and run your analysis

```{r}
#| label: cleaning and mutating personal data
#| include: false

steps <- steps |> #using step data
  mutate(day.of.week = factor(day.of.week,
                              levels = c("M", "T", "W", "Th", "F", "Sa", "Su"))) #making day of week into factors with 7 levels

wvw_steps <- steps |> #creating new weekday vs weekend df
  clean_names() |> #cleaning column names
  mutate(day_of_week = case_when(
    day_of_week %in% c("Sa", "Su") ~ "Weekend", 
    TRUE ~ "Weekday")) #making M-F weekday and S&S weekend
```

### Checking assumptions:

```{r}
#| label: checking assumptions

#checking distribution/normaliy with a histogram
hist(wvw_steps$steps_taken) #normal

#boxplot to check spread and outliers
ggplot(wvw_steps, #using day type df
       aes(x = day_of_week, #x axis
           y = steps_taken))+ #y axis
  geom_boxplot() #normal boxplot

#checking homogeneity of variance
var.test(steps_taken ~ day_of_week, #steps vs day type
         data = wvw_steps) #step data
```

### running the analysis (t-test)

```{r}
#| label: running analysis
#running t-test with step data
t.test(steps_taken~day_of_week, 
       data = wvw_steps)

#calculating effect size with Cohen's d
cohen.d(steps_taken ~ day_of_week,
        data = wvw_steps)
```

## 4d. Create a visualization

```{r}
#| label: creating personal data visualization

#base layer
ggplot(wvw_steps, #using step data
       aes(x = day_of_week, #x axis
                      y = steps_taken, #y axis
                      fill = day_of_week))+ #coloring by day of week type
  geom_boxplot(outlier.shape = 4, #outlier shape (no outliers)
               alpha = 0.7)+ # boxtransparency 
  geom_jitter(width = 0.15, #controlling jitter (how much points can move horizontally)
              size = 2, #point size
              alpha = 0.6, #point transparency
              shape = 21)+ #point shape
  scale_fill_manual(values = c("Weekday" = "dodgerblue", #changing weekday color
                               "Weekend" = "orchid"))+ #changing weekend color
  labs(title = "Step Count by Day Type (Weekday versus Weekend)", #titling plot
       x = "Day Type", #x axis
       y = "Step Count (steps)")+ #y axis
  theme_minimal()+ #changing theme
  theme(legend.position = "none")+ #removing legend
  stat_summary(fun = mean, #adds mean of each y value for each x group
               geom = "point", #adds mean as a point
               size = 4, #size of point
               color = "hotpink") #color of point
  

```

## 4e. Write a caption

**Figure 2. Mean daily step count is greater during weekdays versus weekends.** The boxplots show the differences in daily step count between weekdays and weekends. They show the distribution of daily step counts for weekdays and weekends, with the boxes representing the interquartile range (blue = weekday, purple = weekend) – including the upper and lower quartiles as well as the minimums and maximums – and the horizontal line indicating the median. Individual observations are overlaid as jittered points to display the underlying data. The pink points are the mean step count for each group.

## 4f. Write about your results

Daily step count differed significantly between weekdays and weekend, with more activity (steps) on weekdays versus weekends. On average, my step counts were substantially higher on weekdays (mean = 6749 steps) compared to weekends (mean = 2527 steps). The difference was statistically significant (Welch's two-sample t-test, t(26.789) = 7.45, 95% CI: \[3060.547, 5384.587\], p \< 0.01, $\alpha$ = 0.05) and associated with a large effect size (Cohen's d = 2.23), indicating a strong difference in activity levels between day types.

## 4g. Sharing your affective visualization

\~ donee :) \~
